---
title: "RMS Titanic"
author: "Oliver Buchwald"
date: 2025-02-05
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class
-   Sex
-   Age
-   Survived
-   n

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
-   Are there any differences?
    -   Wikipedia estimates that there were 2224 people aboard
    -   According to our dataset, there were 2201 people onboard
-   If yes, what might account for those differences?
    -   This is a difference of 23 people. Most likely, our dataset and Wikipedia are referencing different records. It's possible there are people who bought tickets, but didn't end up making it on board. Or there could be stowaways on board the ship. Or passengers who died but were not found, and didn't have relatives or a record of their voyage.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survived_q3 <-
  df_titanic %>% 
  filter(Survived == 'Yes')

survived_q3 %>% 
  ggplot(aes(Class, n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Among passengers of 1st and 2nd class, significantly more women than men survived the sinking. In 3rd class there were also more female than male survivors, but by a substantially smaller margin.
-   Crew members are the only exception to this, where more male crewmembers survived than any other demographic. Its possible that this deviation from the trend is due to there being more male than female crew on the ship.
-   A larger number of 1st class passengers survived, compared to 2nd and 3rd classes. 2nd class had the fewest survivors of any group.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}

df_prop %>% 
  filter(Survived == 'Yes') %>%
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Write your observations here.
    -   This feels like the data I wanted from the previous plot!
    -   Proportionally more women survived than men, regardless of class or crew status.
    -   It seems that there is a relationship between higher class and higher proportion of survivors, but I'm not sure if I trust the data to be displaying that accurately.
-   Is there anything *fishy* going on in your plot?
    -   Its really suspicious that all of the proportions of survivors are 1.0 for 1st and 2nd class. That's definitely not accurate to what actually happened on the Titanic.
    -   This is possibly due to not accounting for children separately.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == 'Yes') %>% 
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_grid(rows = 'Age')
```

**Observations**:

-   This data seems much more predictable than past plots.
-   All 1st and 2nd class children survived. In third class, a higher proportion of male children than adults survived, but for women it was roughly equal.
-   Male adults had the lowest survival rate of any group. Where the first graph showed the highest number of male crewmembers surviving, the actual proportion is less than 1/4.
-   Women had relatively high survival rates. Nearly all 1st class and crew women survived, but survival rate among women was progressively lower for lower classes. Interstingly, this trend of lower survival in lower classes didn't hold true for men, where 1st class had the highest proportion of male survivors, followed by 3rd followed by 2nd.
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   Separating out adults and children, the beavior from q4 makes more sense. The survival percentage for upper-class children was very high, but lower for adults, particularly adult men. It appears that the q4 plot only showed the proportion of surviving children, among all classes but crew, which is a super misleading sample.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
