---
title: "Aluminum Data"
author: "Oliver Buchwald"
date: 2025-02-20
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggforce)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`

df_stang_long <- 
  df_stang %>%
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    cols = c(-thick, -alloy)
  ) %>% 
  filter(E > 0) %>% 
  transform(angle = as.integer(angle))


df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
colnames(df_stang_long)

## count number of unique alloys
df_stang_long %>%
  count(alloy)

## count number of samples of each angle
df_stang_long %>%
  count(angle)

## list and count number of samples of each thickness
df_stang_long %>%
  count(thick)
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   Broadly speaking, no - E and mu both have a bit of variation from sample to sample. Properties can very based on material imperfections, varying test setups, and more, so it follows that the data would have some variation from sample to sample.
-   How many aluminum alloys are in this dataset? How do you know?
    -   Only one - All the samples in the data set are listed as al_24st alloy
-   What angles were tested?
    -   0 degrees, 45 degrees, and 90 degrees
-   What thicknesses were tested?
    -   The thicknesses tested were 0.022, 0.032, 0.064 and 0.081. Presumably, these measurements are in inches and correspond to 23, 20,14, and 12 gauge sheet metal, respectively. 0.022 is an odd thickness to have (who uses 23 gauge sheet metal?), so I'm quite curious as to where and why these are the samples at play
-   My question: Does rolling angle have an impact on material properties?

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q1 here
##convert angle back to categorical data
df_stang_plot <-
  df_stang_long %>%
  transform(angle = as.character(angle))

df_stang_plot %>%
  ggplot(aes(x = angle, y = E, fill = angle)) +
  geom_boxplot() + 
  labs(title = 'Young’s Modulus Relative to Rolling Angle', 
       x = 'Rolling Angle (degrees)', 
       y = 'Young’s Modulus (E)')

df_stang_plot %>%
  ggplot(aes(x = angle, y = E, color = angle)) +
  geom_jitter(size = 3) +  
  labs(title = 'Young’s Modulus Relative to Rolling Angle', 
       x = 'Rolling Angle (degrees)', 
       y = 'Young’s Modulus (E)') 
```

**Observations**:

-   There does not seem to be a strong trend here. It appears from the boxplot that the 45 degree samples are more consistent around the median value, which is slightly higher than the samples rolled at other angles. However, the maximum and minimum values are pretty similar regardless of direction, which makes me suspicious of drawing any conclusion. The scatterplot further confirms my suspicion that there doesn't really seem to be any meaningful trend.

```{r q3-task-pt2}

df_stang_plot %>%
  ggplot(aes(x = angle, y = nu, fill = angle)) +
  geom_boxplot() + 
  labs(title = "Poisson's Ratio Relative to Rolling Angle", 
       x = 'Rolling Angle (degrees)', 
       y = "Poisson's Ratio (nu)")

df_stang_plot %>%
  ggplot(aes(x = angle, y = nu, color = angle)) +
  geom_jitter(size = 3) + 
  labs(title = "Poisson's Ratio Relative to Rolling Angle", 
       x = 'Rolling Angle (degrees)', 
       y = "Poisson's Ratio (nu)") 

df_stang_plot %>%
  ggplot(aes(x = nu, color = angle)) +
  geom_density() + 
  labs(title = "Density of Poisson's Ratio", 
       x = "Poisson's Ratio (nu)") 
```

**Observations (Part 2!)**:

-    The boxplot for Poisson's ratio shows a median which is comparatively higher at 45 degrees than other angles. From a mechanical perspective, this makes some degree of sense, since Poisson's ratio relates strain in perpendicular directions, and rotating some microstructure by 45 degrees could allow a better path for that to happen.
-   Comparing to the scatterplot, this interpretation seems a bit more substantiated than for Young's modulus. The data does seem to show a concentration in 45-degree measurements around 0.330 that is much more defined than anything for the 0 or 90 degree samples. However, it still has a fairly high number of outliers (3/8 samples), and I'm hesitant to make any conclusion off of so few samples.
-   The density plot shows a more defined mode for the 45 degree parts, which is another visualization for what I'm getting at, but it feels frankly kind of silly to make a density plot of so few points.
-   All of this to say: I don't think its reasonable to conclude a relationship from this data. Its possible that more investigation, and particularly testing more samples *might* beget some trend, but it just as well might not. If there is orientation-related variation in strain, it seems to be very small, on the order of 1.8% (if we compare the medians). Hopefully, anyone who's designing parts with these materials has a factor of safety of more than 1.01, so its likely that this investigation isn't needed.

### **q4** Consider the following statement:

> "A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%
  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   It seems to contradict the claim. Greater thickness appears to correspond to a lower elasticity modulus, at least for the 0.081 thickness
-   Is this evidence *conclusive* one way or another?
    -   No
    -   Outside of the 0.081 thickness, the other three thicknesses are pretty much comparable. If we exclude that data, the lowest value for E is found at the 0.032 thickness, the highest is shared by 0.022 and 0.064 thickness, but the medians and ranges really show no discernible trend.
    -   This graph is also not a clear answer to that question, in my opinion. Why are we considering nu? Poisson's ratio should be a dimensionless constant for a material, so I suppose it could be used as a control against material chages, but this graph is super odd to me. I'd consider making a box or scatter plot of this data instead, bucketed by material thickness. Or maybe a geom_smooth plot overlaid with a scatter plot, to try and get an idea of the possible trend based on size. I feel like I've made enough graphs for today, though.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
