---
title: "Gapminder"
author: "Oliver Buchwald"
date: 2025-03-02
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder %>%
  colnames()
gapminder %>%
  size_sum()
```

**Observations**:

-   Write all variable names here
    -   country
    -   continent
    -   year
    -   life expectancy
    -   population
    -   GDP per capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`

year_max <-
  gapminder %>%
  summarize(year = max(year)) %>%
  pull(year) 
year_min <- 
  gapminder %>%
  summarize(year = min(year)) %>% 
  pull(year) 
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

df_q2 <-
  gapminder  %>%
  filter(year == min(year))

df_q2 %>% 
  ggplot(aes(gdpPercap, continent, fill = continent)) +
  geom_boxplot(
    data = . %>%
      filter(continent != "Oceania")
  ) + 
  geom_jitter(
    data = . %>%
      filter(continent == "Oceania")
  ) +
  scale_x_log10() + 
  theme(legend.position = "none")

```

**Observations**:

-   Africa has he lowest GDP per capita
-   Asia has a very large outlier, and the widest range of GDPs
-   Oceania has only 2 data points, both of which are quite high compared to the glove as a whole
-   The Americas are merged into one category in this data set. I'm curious if there would be any distinctive split between North & South America if they were plotted separately

**Difficulties & Approaches**:

-   The extreme outlier with a high GDP in asia, which makes the data quite hard to view - I'm using a log scale to correct against this
-   I considered a scatter plot instead of the box plot, but there are enough data points in this set that it was unclear to see trends. Perhaps a sinaplot would help with this, but I didn't feel it was quite worth it at this stage
-   The log scale does make comparisons of the amount of variation in each continent difficult. It looks like Europe and Africa have an equally wide distribution, but my next plot will show this is not the case.

```{r q2-nonlog plot}
df_q2 %>% 
  ggplot(aes(gdpPercap, continent, fill = continent)) +
  geom_boxplot() +
  theme(legend.position = "none") +
  xlim(0,15000)
```

**Observations**:

-   This plot is not complete! I set limits on the x axis so that it excludes the very high outlier in Asia, and adjusted it back to a regular, non-log scale
-   Excluding that datapoint makes it clearer to see the values of outliers for other continents. This plot exists primarily to validate my filtering on the next question.
-   These axes really drive home that Africa has a lower GDP than other continents, and emphasizes that there are only a few outliers in Asia among an otherwise low GDP
-   Europe appears to have a larger standard deviation than other countries. While its min-max range is smaller than Asia, it has much more variation within that range.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
outlier_list <-
  df_q2 %>%
    group_by(continent) %>%
      filter(
        gdpPercap < mean(gdpPercap) - 2*sd(gdpPercap)| 
        gdpPercap > mean(gdpPercap) + 2*sd(gdpPercap)
      ) %>%
    select(country, continent, gdpPercap, year) 
  
outlier_list


outlier_list %>%
  pull(country)

```

**Observations**:

-   Identify the outlier countries from q2
    -   First I tried filtering by percentile, I quickly realized that isn't what we need here- it just finds the max and min values of the dataset, and switched to using standard deviation to filter out all data points which are \>2 std dev from the mean
    -   The table above lists 7 countries which could be considered outliers based on that metric. This is corroborated by the 2nd boxplot from the previous question, which confirms that the GDP listed represents the actual outliers in the dataset
        -   The countries in question are Angola, Canada, Gabon, Kuwait, South Africa, Switzerland, and the United States
    -   I would also consider the continent of Oceania to be an outlier in this dataset as it only contains two countries. The countries aren't outliers relative to the continent because there's not enough data to say what an outlier is, but they're both very high relative to the entire dataset.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

#refilter outlier_list for both year_min and year_max
outlier_list <-
  gapminder %>%
  filter(year == max(year) | year == min(year)) %>%
    group_by(continent, year) %>%
      filter(
        gdpPercap < mean(gdpPercap) - 3*sd(gdpPercap)| 
        gdpPercap > mean(gdpPercap) + 3*sd(gdpPercap)|
        continent == "Oceania"
      ) %>%
    select(country, continent, gdpPercap, year)

#begin by only using year_min from q2
gapminder %>% 
  filter(year == max(year) | year == min(year)) %>%
  ggplot(aes(continent,gdpPercap)) +
  geom_boxplot(
    data = . %>%
      filter(continent != "Oceania")
  ) +
  geom_point(
    data = outlier_list %>%
      group_by(year),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +  
  facet_wrap(~year) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r q4-skew}
#thanks ZDR for this code chunk
gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  group_by(continent, year) %>%
  summarize(skew = moments::skewness(gdpPercap))
```

**Observations**:

-   There is a substantial amount of change in the 55 year period of this data set. I had to adjust my definition of an outlier to a country that is within 3 standard deviations of the mean, rather than 2 because the countries which were outliers changed from year to year. This resulted in a plot with too many individual color-coded points to be readable.
-   On average, across every continent, GDP increased from 1952 to 2007. There are some individual countries which are exceptions to this, but for the median, 75th percentile, and maximum of each continent there is an upward trend. Kuwait is one noteable exception to this trend - Its GDP per capita dropped by nearly 50%. Africa, and the Americas also had the minimum GDP drop. What happened in Kuwait?
-   Most continents developed a less normal distribution in GDPs over the length of the dataset. The boxplots became less symmetric across the mean, with Asia and Africa developing a less negative skew while Europe and the Americas skew more negative.

```{r q4-nonlog}
## A second plot removing the log scale

#begin by only using year_min from q2
gapminder %>% 
  filter(year == max(year) | year == min(year)) %>%
  ggplot(aes(continent,gdpPercap)) +
  geom_boxplot(
    data = . %>%
      filter(continent != "Oceania")
  ) +
  geom_point(
    data = outlier_list %>%
      group_by(year),
    mapping = aes(color = country)
  ) + 
  facet_wrap(~year) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
```

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

**Question:** What is going on with Kuwait? Is there anything else weird about its data?

```{r q5-task1}
## TASK: Your first graph

# Compute mean GDP
gdp_means <- gapminder %>%
  group_by(year) %>%
  summarise(
    mean_gdp_global = mean(gdpPercap, na.rm = TRUE),
    mean_gdp_asia = mean(gdpPercap[continent == "Asia"], na.rm = TRUE),
    kuwait_gdp = gdpPercap[country == "Kuwait"]
  ) %>%
  pivot_longer(cols = c(mean_gdp_global, mean_gdp_asia, kuwait_gdp), 
               names_to = "Metric", 
               values_to = "GDP")


#plot agaisnt kuwait
gdp_means %>%
  ggplot(aes(year, GDP, color = Metric)) +
  geom_line() +
  
  ggtitle("Kuwait GDP Per Capita vs Global & Africa Mean GDP") +
  ylab("GDP Per Capita (log scale)") +
  scale_y_log10()

```

```{r q5-task1-part2}

# compute means of each  parameter for the globe and for asia
metrics_means <- gapminder %>%
  group_by(year) %>%
  summarise(
    mean_gdp_global = mean(gdpPercap, na.rm = TRUE),
    mean_gdp_asia = mean(gdpPercap[continent == "Asia"], na.rm = TRUE),
    kuwait_gdp = gdpPercap[country == "Kuwait"],

    mean_pop_global = mean(pop, na.rm = TRUE),
    mean_pop_asia = mean(pop[continent == "Asia"], na.rm = TRUE),
    kuwait_pop = pop[country == "Kuwait"],

    mean_lifeExp_global = mean(lifeExp, na.rm = TRUE),
    mean_lifeExp_asia = mean(lifeExp[continent == "Asia"], na.rm = TRUE),
    kuwait_lifeExp = lifeExp[country == "Kuwait"]
  ) %>%
  
  #pivot for better plotting
  pivot_longer(cols = c(mean_gdp_global, mean_gdp_asia, kuwait_gdp, 
                        mean_pop_global, mean_pop_asia, kuwait_pop, 
                        mean_lifeExp_global, mean_lifeExp_asia, kuwait_lifeExp), 
               names_to = "Metric", values_to = "Value") %>%
  
  #add location category to color code
  mutate(Location = case_when(
    str_detect(Metric, "kuwait") ~ "Kuwait",
    str_detect(Metric, "asia") ~ "Asia",
    str_detect(Metric, "global") ~ "Global"
  ),
  # add metric label for faceting
  Metric = case_when(
    str_detect(Metric, "gdp") ~ "GDP Per Capita",
    str_detect(Metric, "pop") ~ "Population",
    str_detect(Metric, "lifeExp") ~ "Life Expectancy"
  ))

# Plot all three metrics
  metrics_means %>%
  ggplot(aes(x = year, y = Value, color = Location)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~Metric, scales = "free_y") + 
  ggtitle("Kuwait vs Global & Asia Means") +
  ylab("Value") +
  scale_y_log10() + 
  theme(
    legend.title = element_blank(), 
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

-   Kuwait's GDP began to drop drastically from 1970 to 1980. In the 90s and 2000s it began increasing again. It is still significantly higher than the average GDP of both the continent and the globe.
    -   This seems to correspond to the 1973 oil crisis, as Kuwait's economy is significantly reliant on oil
-   It seems that the mean of GDPs in Asia experiences a slight dip in that time period as well - possibly because Kuwait's GDP is just that high, and its crash dragged down the mean of the continent as a whole.
-   Kuwait has a generally high life expectancy and generally low population, but the turmoil in its GDP did not seem to have an impact on either of these values. Population and life expectancy continually trended upwards

**Question:** Is there a relationship between GDP, Life Expectancy & population?

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  filter(year == year_max | year == year_min) %>%
    ggplot(aes(x = lifeExp, y = gdpPercap, size = pop, color = continent)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(1,8)) +
    ylim(0,50000) +
    ggtitle(paste("GDP vs Life Expectancy")) +
    facet_wrap(~year) +
    xlab("Life Expectancy") +
    ylab("GDP Per Capita") +
    guides(
        size = guide_legend(title = "Population (Size)"),
        color = guide_legend(title = "Continent (Color)")
      )
```

-   Note: the y axis is truncated to exclude Kuwait from the 1952 data so that the plots are clearer to read. It has a GDP of 108,000 per capita, a life expectancy of 55.5, and a population of 160,000
-   Generally speaking, high GDP corresponds to higher life expectancy.
-   There is not a strong relationship between population and either GDP or life expectancy. It appears that the countries with the highest population have a life expectancy that is near the median of the dataset. It effectively seems like countries with a low life expectancy do not have a large population, but a high life expectancy is not a guarantee of high population - likely due to birthrates
-   European countries tend to have the highest life expectancy and GDPs, while African countries are the lowest.
-   The data is more heavily segmented by continent in 2007 and 1952. On a global scale, the data from 1952 shows values for countries in each continent that are much closer together than 2007.
-   Life expectancy overall increased from 1952 to 2007 by around 10 years.

**Question:** How has population growth happened over time?

```{r q5-task3}
## TASK: Your third graph

#normalize population growth
df_pop_growth <-
  gapminder %>%
  group_by(country) %>%
  mutate(pop_normalized = pop / pop[year == year_min]) %>%
  ungroup %>%
  mutate(legend_label = paste0(continent, " - ", country ))



# plot all countries
df_pop_growth %>% 
  ggplot(aes(year, pop_normalized, group = country)) +
  
  #plot all countries in transparent lines
  geom_line(alpha = 0.2) +
  #geom_jitter(alpha = 0.2) +
  
  #separately plot max and min country on each continent in color
  geom_line( 
    data = . %>%
      filter(year == year_max) %>%
      group_by(continent) %>%
      filter(
        pop_normalized == max(pop_normalized) |
        pop_normalized == min(pop_normalized)
      ) %>%
      select(country) %>%
      group_by(continent) %>%
      inner_join(df_pop_growth, by = c("continent","country")) ,
    aes(color = legend_label),
    linewidth = 1.2
  ) + 
  
  #add smooth trendline for each continent
  geom_smooth(aes(group = continent), color = "red", alpha = 0.5, linewidth = 0.8, method = "lm", se = FALSE) +
  
  ggtitle("Population Growth Over Time") +
  ylab("Population Growth (normalized to 1955)") +
  facet_wrap(~continent) +
  guides(color = guide_legend(ncol = 2)) +
  theme(
    legend.text = element_text(size = 6), 
    legend.title = element_blank(),
    legend.position.inside = c(1.03,0),
    legend.justification = c(1, 0),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

-   For these plots, I normalized the population growth by dividing population of each country by its population in 1955. This makes it much clearer to compare growth without a few high-population outliers compressing the axes excessively.
-   Asia and africa have the greatest proportion of population growth rate of any continent
-   Interestingly, Kuwait has the highest rate of population increase per capita, despite its low population. This is a noteable limitation of this plot that it doesn't consider actual population size, only growth rates. Djibouti and another country in Asia are other noteably high-growth countries, with the population growing 8x in this 55year period
-   Europe and Oceania both experience very low growth in population, with a growth rate of \~1.5x and 2x respectively. Bulgaria is the country with the lowest population growth of all, at a rate of \<1% growth over the entire dataset.
