---
title: "Gapminder"
author: "(Your name here)"
date: 2025-03-03
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder %>%
  colnames()
gapminder %>%
  size_sum()
```

**Observations**:

-   Write all variable names here
    -   country
    -   continent
    -   year
    -   life expectancy
    -   population
    -   GDP per capita

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`

year_max <-
  gapminder %>%
  summarize(year = max(year)) %>%
  pull(year) 
year_min <- 
  gapminder %>%
  summarize(year = min(year)) %>% 
  pull(year) 
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

df_q2 <-
  gapminder  %>%
  filter(year == min(year))

df_q2 %>% 
  ggplot(aes(gdpPercap, continent, fill = continent)) +
  geom_boxplot(
    data = . %>%
      filter(continent != "Oceania")
  ) + geom_jitter(
    data = . %>%
      filter(continent == "Oceania")
  ) +
  scale_x_log10() + 
  theme(legend.position = "none")

```

**Observations**:

-   I initially used

**Difficulties & Approaches**:

-   Write your challenges and your approach to solving them
-   The extreme outlier with a high GDP in asia, which makes the data quite hard to view - I'm using a log scale to correct against this
-   I considered a scatter plot instead of the box plot, but there are enough data points in this set that it was unclear to see trends. Perhaps a sinaplot would help with this, but I didn't feel it was quite worth it at this stage

```{r q2-nonlog plot}
df_q2 %>% 
  ggplot(aes(gdpPercap, continent, fill = continent)) +
  geom_boxplot() +
  theme(legend.position = "none") +
  xlim(0,15000)
```

**Observations**:

-   This plot is not complete! I set limits on the x axis so that it excludes the very high outlier in Asia, and adjusted it back to a regular, non-log scale
-   Excluding that datapoint makes it clearer to see the values of outliers for other continents. This plot exists primarily to validate my filtering on the next question.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
outlier_list <-
  df_q2 %>%
    group_by(continent) %>%
      filter(
        gdpPercap < mean(gdpPercap) - 2*sd(gdpPercap)| 
        gdpPercap > mean(gdpPercap) + 2*sd(gdpPercap)
      ) %>%
    select(country, continent, gdpPercap) 
  
outlier_list


outlier_list %>%
  pull(country)

```

**Observations**:

-   Identify the outlier countries from q2
    -   First I tried filtering by percentile, I quickly realized that isn't what we need here- it just finds the max and min values of the dataset, and switched to using standard deviation to filter out all data points which are \>2 std dev from the mean
    -   The table above lists 7 countries which could be considered outliers based on that metric. This is corroborated by the 2nd boxplot from the previous question, which confirms that the GDP listed represents the actual outliers in the dataset
    -   I would also consider the continent of Oceania to be an outlier in this dataset as it only contains two countries. The countries aren't outliers relative to the continent because there's not enough data to say what an outlier is, but they're both very high relative to the entire dataset.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent

#begin by only using year_min from q2
df_q2 %>% 
  ggplot(aes(gdpPercap, continent, fill = continent)) +
  geom_boxplot(
    data = . %>%
      filter(continent != "Oceania")
  ) + geom_jitter(
    data = . %>%
      filter(continent == "Oceania")
  ) +
  scale_x_log10() + 
  theme(legend.position = "none")
  
```

**Observations**:

-   Write your observations here

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
```

-   (Your notes and observations here)

```{r q5-task2}
## TASK: Your second graph
```

-   (Your notes and observations here)

```{r q5-task3}
## TASK: Your third graph
```

-   (Your notes and observations here)
