---
title: "covid-births"
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## c06 Data pulls

```{r pull-census}
## TASK: Load the census bureau data with the following tibble name.


pop <- "./data/ACSDT5Y2018.B01003-Data.csv"
pop_columns <- "./data/ACSDT5Y2018.B01003-Column-Metadata.csv"

df_pop <- read_csv(pop, skip = 2, 
                    na = "*****", 
                    col_types = "ccd_d__", 
                    col_names = c("id", 
                                  "Geographic Area Name", 
                                  "Estimate!!Total", 
                                  "Margin of Error!!Total"))
head(df_pop)
```

```{r find-NYT}
## TASK: Find the URL for the NYT covid-19 county-level data
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"

```

```{r download-NYT}
## NOTE: No need to change this; just execute
## Set the filename of the data to download
filename_nyt <- "./data/nyt_counties.csv"

## Download the data locally
curl::curl_download(
        url_counties,
        destfile = filename_nyt
      )

## Loads the downloaded csv
df_covid <- read_csv(filename_nyt)
head(df_covid)
```

### Clean & Filter data

```{r census-fips}
## TASK: Create a `fips` column by extracting the county code
df_q3 <- 
  df_pop %>%
  mutate(fips = str_extract(id, "[^US]*$")) %>%
  select(-id) %>%
  select(fips, 'Geographic Area Name', 'Estimate!!Total', 'Margin of Error!!Total')

df_q3
```

```{r q4-task}
## TASK: Join df_covid and df_q3 by fips.
df_q4 <- left_join(df_covid, df_q3, by = "fips") # %>%
  # filter(!is.na(fips)) #remove NYT data that is missing a fips code
   
```

sort organized data

```{r rename}
## NOTE: No need to change; run this to produce a more convenient tibble
df_data <-
  df_q4 %>%
  select(
    date,
    county,
    state,
    fips,
    cases,
    deaths,
    population = `Estimate!!Total`
  )

```

# A

## Covid Birth Comparison

Pull CDC wonder data from here: <https://wonder.cdc.gov/natality-expanded-current.html>

This code processes an excel file downloaded by using this tool to generate a list of birth rates grouped by county and month

```{r pull-births}

## load dataset
library(readxl)
birth <- "./data/covid_birth_data.xlsx"
df_births <- read_excel(birth, sheet = 1)

## tidy data
df_births_cleaned <- 
  ## remove extra rows for computed totals
  df_births[!apply(df_births, 1, function(row) any(grepl("Total", row, ignore.case = TRUE))), ] %>%
  ## simplify colnames for consistency
  rename(births = Births) %>%
  rename(fips = 'County of Residence Code') %>%
  mutate(fips = as.character(fips)) %>%
  mutate(fips = str_pad(as.character(fips), width = 5, pad = "0")) %>%
  ## encode date column as the 1st of the month following the end of the month eg January 2020 -> 2020-02-01
  mutate(date = str_glue("{`Year Code`}-{`Month Code`}-01")) %>%
  mutate(date = as.Date(date)) %>%
  mutate(birth_date = date %m+% months(1)) %>%
  mutate(sex_date = date %m-% months(9)) %>%
  ## pull only relevant data
  select(fips, birth_date, sex_date, births)
  

head(df_births_cleaned)
```

```{r consolidate-covid-data}

df_covid_month <- df_covid %>%
  ## round date to the 1st of next month
  mutate(year_month = ceiling_date(date, unit = "month")) %>% 
  ## pull only the last data point of the month per county
  group_by(fips, year_month) %>%
  filter(date == max(date)) %>%  
  arrange(fips, year_month) %>%  
  ungroup() %>%
  ## compute new cases and deaths occuring that month
  group_by(fips) %>%
  mutate(
    new_cases = cases - lag(cases, default = 0),   
    new_deaths = deaths - lag(deaths, default = 0) 
  ) %>%
  ## log only relevant data
  select(fips, year_month, cases, new_cases,deaths, new_deaths) %>%  
  ungroup()

head(df_covid_month)

```

```{r join-all-data}
df_covid_births <- 
  df_births_cleaned %>%
  left_join(df_covid_month, by = c("fips","sex_date" = "year_month")) %>%
  # rename(sex_date = year_month) %>%
  left_join(df_q3, by = "fips") 
  # filter(!is.na(fips)) #remove NYT data that is missing a fips code
   
head(df_covid_births)
```

## Test Plots

You can also embed plots, for example:

```{r birthgraph}
df_covid_births %>%
  ggplot(aes(x = sex_date, y = births, group = fips)) +
  geom_line(alpha = 0.5) + 
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "red")

df_covid_births %>%
  ggplot(aes(x = sex_date, y = new_cases, group = fips)) +
  geom_line(alpha = 0.5) + 
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "red")



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
