---
title: "Setup: RStudio Shortcuts"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 10
reading: 5
---

# Setup: RStudio Shortcuts

*Purpose*: Your ability to get stuff done is highly dependent on your fluency with your tools. One aspect of fluency is knowing and *using* keyboard shortcuts. In this exercise, we'll go over some of the most important ones.

*Reading*: [Keyboard shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts); [Code Chunk Options](https://rmarkdown.rstudio.com/lesson-3.html)

*Note*: Use this reading to look up answers to the questions below. Rather than *memorizing* this information, I recommend you download a [cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/rstudio-ide.pdf), and either print it out or save it in a convenient place on your computer. Get used to referencing your cheatsheets while doing data science---practice makes perfect! - If this link is broken, try searching "Rstudio IDE cheatsheet" on the web.

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### **q1** What do the following keyboard shortcuts do?

-   Within the script editor or a chunk

    -   `Alt` + `-`

        -   draws a little arrow

    -   `Shift` + `Cmd/Ctrl` + `M`

        -   does this %\>%

    -   `Cmd/Ctrl` + `Enter`

        -   runs code

    -   `F1` (Note: on a Mac you need to press `fn` + `F1`)

        -   mutes my speakers?

        -   can get function help page if a function is selected

    -   `Cmd/Ctrl` + `Shift` + `C`

        -   comment

-   Within R Markdown

    -   `Cmd/Ctrl` + `Alt` + `I`
        -   new code chunk

-   Within a chunk

    -   `Shift` + `Cmd/Ctrl` + `Enter`

        -   runs code - same as ctrl+enter?

    -   `Ctrl` + `I` (`Cmd` + `I`)

        -   unindents

Try this below!

```{r}
## Re-indent these lines
c(
  "foo",
  "bar",
  "goo",
  "gah"
)
```

### **q2** For a chunk, what header option do you use to

-   Run the code, don't display it, but show its results?

    ```{r echo = FALSE}
    ```

-   Run the code, but don't display it or its results?

    ```{r include = FALSE}
    ```

### **q3** How do stop the code in a chunk from running once it has started?

Click the stop button on the chunk

### **q4** How do you show the "Document Outline" in RStudio?

*Hint*: Try googling "rstudio document outline"

ctrl+shift+o

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup04-rstudio-shortcuts-assignment.Rmd).
